---
title: "A socio-economic Model of Residential Segregation, Neighborhood Change and Housing Inequality"
subtitle: "Part IIIf: Testable Hypotheses"
author: "Malte Grönemann"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, 
                      message = FALSE, 
                      warning = FALSE, 
                      error = FALSE,
                      out.width = "50%")

library(nanoparquet)
library(dplyr)
library(tidyr)
library(broom)
library(Hmisc)
library(ggplot2)
library(ggpubr)
library(ggcorrplot)
library(fixest)
library(modelsummary)

#data_directory <- "./data/hypotheses/" # TODO change directory
data_directory <- "./data/" # TODO change directory

unit_data <- read_parquet(paste(data_directory, "unit_data.parquet", sep = ""))
nb_data <- read_parquet(paste(data_directory, "nb_data.parquet", sep = ""))
city_data <- read_parquet(paste(data_directory, "city_data.parquet", sep = ""))

# create ad merge average Gini values for the three distribution shapes
gini_df <- unit_data %>%
  group_by(distribution) %>%
  summarise(dist_gini = Gini(hh_income, na.rm = TRUE))

unit_data <- unit_data %>%
  full_join(gini_df, by = "distribution")

nb_data <- nb_data %>%
  full_join(gini_df, by = "distribution")

city_data <- city_data %>%
  full_join(gini_df, by = "distribution")
```

# Experiment Description

One of the main motivations of the new theory is to link multiple phenomena in housing markets, most notably residential segregation, residential mobility, and housing inequality. As such, this mdel is able to produce new hypotheses about the relationships between these phenomena. This analysis derives a (likely incomplete) list of these hypotheses.

As the main argument is that both households and landlords are spatially interdependent in their behavior, segregation is an important macro-context in which they act. E.g., when neighborhood averages are more distinct from each other, the variables derived from the averages will be more unequal. Most hypotheses are based on this property of the model, making residential segregation a moderator for individual-level relationships. This choice also does not require to empirically estimate the model parameters, especially the relative importance of housing and neighborhood quality, which are difficult to measure.


## Experiment Description

The main experiment and the robustness checks use global parameters that span the entire range possible. However, some values of these parameters are very unlikely to be realised in the real world. I therefore simulate a data set for hypothesis generation that only slightly varies the global parameters around realistic values. As households tend to value housing quality to a greater extent than neighbourhood quality, I use values of 0.2, 0.3 and 0.4 for the relative importance parameter $a$ in the utility function. Income and status also tend to be highly correlated, although not perfectly. I use correlation parameter values of 0.4, 0.5 and 0.6 (the correlation parameter tends to produce Pearson correlations between the two variables higher than the parameter value). And housing quality decays slowly over time, so I use decay parameter values of 0.95, 0.925 and 0.9. I also vary the level of income inequality in three levels (distribution parameters of 1.5, 2 and 2.5), corresponding to Gini coefficients of 0.36, 0.33, and 0.29.

This selection of parameter values results in 81 combinations and each combination is run ten times, resulting in 810 independent simulation runs of 200 steps. With a grid size of 30 x 30, the raw output data amounts to 145.8 million observations of housing units and 72.9 million of these are analysed.


## List of Hypotheses





# Segregation Levels as Dependent Variable

```{r}
seg_long <- city_data %>%
  select(-ends_with("gini"), -vacancy_seg, dist_gini) %>%
  pivot_longer(cols = ends_with("seg"),
               cols_vary = "slowest",
               names_to = "dimension",
               values_to = "segregation") %>%
  mutate(dimension = if_else(dimension == "inc_seg", "Income", dimension),
         dimension = if_else(dimension == "status_seg", "Status", dimension),
         dimension = if_else(dimension == "quality_seg", "Quality", dimension),
         dimension = if_else(dimension == "rent_seg", "Rent", dimension))

seg_long %>%
  group_by(a_preferences, r_correlation, dimension, dist_gini) %>%
  summarise(segregation = mean(segregation)) %>%
  ggplot() +
  aes(x = dist_gini,
      y = segregation,
      colour = dimension,
      group = dimension) +
  geom_point() +
  geom_line() +
  facet_grid(rows = vars(r_correlation),
             cols = vars(a_preferences), 
             labeller = label_both) +
  labs(title = "Segregation", y = "H^R", 
       x = "Gini of the Income Distribution") +
  theme_bw() +
  theme(legend.position = "bottom")
```


- Neighborhoods are more homogenous in housing quality and rent than in income and status.
- Cities are more segregated, the more income and status are correlated.
- Cities are more segregated, the more housing quality matters to households.
- Correlation and preferences interact in their relation to segregation.


# City-level Inequality as Dependent Variable



```{r}
modelsummary(list(
  feols(data = city_data,
              quality_gini ~ rent_seg + dist_gini,
              cluster = "sample_id"),
  feols(data = city_data,
              quality_gini ~ rent_seg * dist_gini,
              cluster = "sample_id")
),
  output = "html",
  stars = c('**' = .01, '***' = .001), 
  gof_map = c("nobs", "r.squared", "bic"))

# TODO add rent 
```

- More segregated cities have higher inequalities in housing quality and rent.
- Cities with higher income inequality have higher inequalities in housing quality and rent.
- The positive effect of income inequality on housing/rent inequality is larger in more segregated cities (positive interaction effect).

# Relationship between Inequality and Segregation?

```{r}
city_data %>%
  select(ends_with("gini"), ends_with("seg")) %>%
  cor() %>%
  ggcorrplot(method = "circle", type = "upper", 
             show.legend = F, lab = T) +
  labs(title = "Correlation Matrix")
```



# Neighborhood Averages and their Distributions as DV

```{r}
nb_reg <- nb_data %>%
  group_by(sample_id, a_preferences, r_correlation, d_decay, dist_gini, t) %>%
  summarise(nb_gini_inc = Gini(ave_inc),
            nb_gini_quality = Gini(ave_quality))

nb_reg <- city_data %>%
  select(sample_id, inc_seg, rent_seg) %>%
  full_join(nb_reg, by = "sample_id")

modelsummary(list(
  feols(data = nb_reg,
              nb_gini_inc ~ inc_seg + dist_gini,
              cluster = "sample_id"),
  feols(data = nb_reg,
              nb_gini_inc ~ inc_seg * dist_gini,
              cluster = "sample_id")
),
  output = "html",
  stars = c('**' = .01, '***' = .001), 
  gof_map = c("nobs", "r.squared", "bic"))
```

```{r}
modelsummary(list(
  feols(data = nb_reg,
              nb_gini_quality ~ rent_seg + dist_gini,
              cluster = "sample_id"),
  feols(data = nb_reg,
              nb_gini_quality ~ rent_seg * dist_gini,
              cluster = "sample_id")
),
  output = "html",
  stars = c('**' = .01, '***' = .001), 
  gof_map = c("nobs", "r.squared", "bic"))
```

higher inequality in incomes between nb is equivalent to income segregation. but segregation refers to between- and within variance, not the values of the averages.


- The positive effect of income inequality on average neighborhood income/housing/rent inequality is larger in more segregated cities.


# Hypotheses related to Neighborhood Stability and Change

```{r}
stability_data <- nb_data %>%
  group_by(sample_id, nb) %>%
  summarise(sd_rank_inc = sd(rank_inc),
            sd_rank_rent = sd(rank_rent),
            sd_inc_t = sd(ave_inc, na.rm = TRUE),
            sd_rent_t = sd(ave_rent, na.rm = TRUE)) %>%
  group_by(sample_id) %>%
  summarise(ave_sd_rank_inc = mean(sd_rank_inc),
            ave_sd_rank_rent = mean(sd_rank_rent),
            ave_sd_inc = mean(sd_inc_t),
            ave_sd_rent = mean(sd_rent_t))

stability_data <- city_data %>%
  group_by(sample_id, a_preferences, r_correlation, d_decay, dist_gini) %>%
  summarise(ave_seg_inc = mean(inc_seg),
            ave_seg_rent = mean(rent_seg)) %>%
  full_join(stability_data, by = "sample_id")

modelsummary(list(
  feols(data = stability_data,
              ave_sd_rank_inc ~ ave_seg_inc,
              cluster = "sample_id"),
  feols(data = stability_data,
              ave_sd_inc ~ ave_seg_inc,
              cluster = "sample_id"),
  feols(data = stability_data,
              ave_sd_rank_rent ~ ave_seg_rent,
              cluster = "sample_id"),
  feols(data = stability_data,
              ave_sd_rent ~ ave_seg_rent,
              cluster = "sample_id")
),
  output = "html",
  stars = c('**' = .01, '***' = .001), 
  gof_map = c("nobs", "r.squared", "bic"))
```

The regression shows that (both for ranks (1+3) and neighborhood averages(2+4)), higher segregation has a negative effect on average standard deviation in neighborhood income and rent.

- Neighborhoods are more stable over time in more segregated cities.

in bounded rationality models: because absolute difference are greater with greater inequality, neighborhoods should be more stable in equilibrium.


# Hypotheses related to Vacancies

- The more segregated a city by income, the more concentrated are vacancies in the poorest and most rundown neighborhoods.
- The more segregated a city by income, the lower is the average housing quality and rent of the vacant units.


# Individual-level Hypotheses

- The more segregated a city, the less likely are rich households to move.

- There is a decrease in marginal return of housing quality by income.
- The positive relationship between income and rent is stronger in more segregated cities.

- The more important neighbors status is to households, the lower the housing quality of low status residents even when controlling for income.

