---
title: "A socio-economic Model of Residential Segregation, Neighborhood Change and Housing Inequality"
subtitle: "Part IIId: Analysis of the Vision Experiment"
author: "Malte Grönemann"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, 
                      message = FALSE, 
                      warning = FALSE, 
                      error = FALSE,
                      out.width = "50%")

library(nanoparquet)
library(dplyr)
library(tidyr)
library(Hmisc)
library(ggplot2)

data_directory <- "./data/vision/" # TODO change directory

city_data <- read_parquet(paste(data_directory, "city_data.parquet", sep = ""))
nb_data <- read_parquet(paste(data_directory, "nb_data.parquet", sep = ""))
unit_data <- read_parquet(paste(data_directory, "unit_data.parquet", sep = ""))
```

# Experiment Description

## Experiment Description

In their paper, Laurie and Jaggi (2003) have established that in categorical segregation models, it makes a difference what agents consider their neighborhood. Athough their claim that depending on vision, segregation might not occur at all, was shown to be a specificity of their simulation setup (Fossett and Waren, 2005), they additionally find that segregation tends form larger clusters when vision increases. This lends itself also to look at the scale of segregation, which has been of some interest to scholars (see references). 

Before, I have used a Moore neighborhood of distance 1, which refers to the 8 bordering (including diagonally) cells. Using the vision parameter, I can increase this distance. I vary it between 1 (8 neighbors), 2 (24 neighbors) and 3 (48 neighbors). The model simulates a grid of 30x30 and uses 3 x 3 x 2 = 18 levels of the initial parameters and additionally the 3 levels of vision just discussed. The model repeats every combination 10 times, so the experiment consists of 18 x 3 x 10 = 540 runs of 200 steps. The output results in 30 x 30 x 540 x 200 = 97.2 million observations of housing units. The first 100 observations are deletd as burn-in.

I will follow the implementation of Reardon et al (2008) which create overlapping ego-neighborhoods of varying radii around the focal household to measure segregation at different scales. Existing studies on the scale of segregation typically considered racial/ethnic segregation. To my knowledge, only Reardon and Bischoff (2011) have analysed income segregation at different scales. They find that "upper-income households are much more segregated at a large geographic scale than lower-income households. For high-income households, 60% or more of segregation patterns are due to macro-scale segregation, presumably the concentration of high-income households in wealthy suburban and exurban areas. For low-income households, 40% or less of segregation patterns are due to macro-scale segregation; this implies that the poor are less concentrated spatially than the wealthy in most metropolitan areas" (p. 1125). They also find that the effect of income inequality and income segregation is almost exclusively driven by the increasing macro-scale segregation of affluent households.


## Results Summary




# City Level

## Segregation ($H^R$ Index)


```{r}
seg_long <- city_data %>%
  select(-ends_with("gini")) %>%
  pivot_longer(cols = ends_with("seg"),
               cols_vary = "slowest",
               names_to = "dimension",
               values_to = "segregation") %>%
  mutate(dimension = if_else(dimension == "inc_seg", "Income", dimension),
         dimension = if_else(dimension == "status_seg", "Status", dimension),
         dimension = if_else(dimension == "quality_seg", "Quality", dimension),
         dimension = if_else(dimension == "rent_seg", "Rent", dimension),
         dimension = if_else(dimension == "vacancy_seg", "Vacancy", dimension))

seg_long %>%
  filter(d_decay == 0.95 & dimension != "Vacancy") %>%
  group_by(a_preferences, r_correlation, dimension, vision) %>%
  summarise(segregation = mean(segregation)) %>%
  ggplot() +
  aes(x = vision,
      y = segregation,
      colour = dimension,
      group = dimension) +
  geom_point() +
  geom_line() +
  facet_grid(rows = vars(r_correlation),
             cols = vars(a_preferences), 
             labeller = label_both) +
  labs(title = "Segregation (d_decay == 0.95)", y = "H^R") +
  theme_bw() +
  theme(legend.position = "bottom")

seg_long %>%
  filter(d_decay == 0.8 & dimension != "Vacancy") %>%
  group_by(a_preferences, r_correlation, dimension, vision) %>%
  summarise(segregation = mean(segregation)) %>%
  ggplot() +
  aes(x = vision,
      y = segregation,
      colour = dimension,
      group = dimension) +
  geom_point() +
  geom_line() +
  facet_grid(rows = vars(r_correlation),
             cols = vars(a_preferences), 
             labeller = label_both) +
  labs(title = "Segregation (d_decay == 0.8)", y = "H^R") +
  theme_bw() +
  theme(legend.position = "bottom")
```
## Geographic Scale of Segregation

```{r}

```


## Inequality

```{r}
ggplot(city_data) +
  aes(x = as.factor(vision),
      y = rent_gini,
      fill = as.factor(d_decay)) +
  geom_violin() +
  facet_grid(rows = vars(r_correlation),
             cols = vars(a_preferences), 
             labeller = label_both) +
  labs(title = "Rent Inequality",
       x = "vision", y = "Gini index", fill  = "d_decay") +
  theme_bw() +
  theme(legend.position = "bottom")

ggplot(city_data) +
  aes(x = as.factor(vision),
      y = quality_gini,
      fill = as.factor(d_decay)) +
  geom_violin() +
  facet_grid(rows = vars(r_correlation),
             cols = vars(a_preferences), 
             labeller = label_both) +
  labs(title = "Inequality in Housing Quality",
       x = "vision", y = "Gini index", fill  = "d_decay") +
  theme_bw() +
  theme(legend.position = "bottom")
```

# Neighborhood Level

## Income and Status

```{r}
ggplot(nb_data) +
  aes(y = ave_inc, 
      x = as.factor(vision),
      fill = as.factor(d_decay)) +
  geom_violin() +
  facet_grid(rows = vars(r_correlation),
             cols = vars(a_preferences), 
             labeller = label_both) +
  labs(y = "Average NB Income", fill = "d_decay", x = "turnover") +
  theme_bw() +
  theme(legend.position = "bottom")

ggplot(nb_data) +
  aes(y = ave_status, 
      x = as.factor(vision),
      fill = as.factor(d_decay)) +
  geom_violin() +
  facet_grid(rows = vars(r_correlation),
             cols = vars(a_preferences), 
             labeller = label_both) +
  labs(y = "Average NB Status", fill = "d_decay", x = "turnover") +
  theme_bw() +
  theme(legend.position = "bottom")
```
## Rent and Housing Quality

```{r}
ggplot(nb_data) +
  aes(y = ave_rent, 
      x = as.factor(vision),
      fill = as.factor(d_decay)) +
  geom_violin() +
  facet_grid(rows = vars(r_correlation),
             cols = vars(a_preferences), 
             labeller = label_both) +
  labs(y = "Average NB Rent", fill = "d_decay", x = "turnover") +
  theme_bw() +
  theme(legend.position = "bottom")

ggplot(nb_data) +
  aes(y = ave_quality, 
      x = as.factor(vision),
      fill = as.factor(d_decay)) +
  geom_violin() +
  facet_grid(rows = vars(r_correlation),
             cols = vars(a_preferences), 
             labeller = label_both) +
  labs(y = "Average NB Housing Quality", fill = "d_decay", x = "turnover") +
  theme_bw() +
  theme(legend.position = "bottom")
```

## Neighborhood Stability

When the definition of neighborhood in households utility calculation and rent setting by landlords increases, moves from single households should present less of a shock. It is therefore reasonable to expect that neighborhood stability increases with vision.

```{r}
stability_data <- nb_data %>%
  group_by(sample_id, r_correlation, a_preferences, d_decay, vision, nb) %>%
  summarise(ave_rank_inc = mean(rank_inc),
            sd_rank_inc = sd(rank_inc),
            ave_rank_status = mean(rank_status),
            sd_rank_status = sd(rank_status),
            ave_rank_quality = mean(rank_quality),
            sd_rank_quality = sd(rank_quality),
            ave_rank_rent = mean(rank_rent),
            sd_rank_rent = sd(rank_rent))

ggplot(stability_data) +
  aes(x = ave_rank_inc,
      y = sd_rank_inc,
      shape = as.factor(d_decay),
      linetype = as.factor(d_decay),
      colour = as.factor(vision)) +
  geom_smooth(se = FALSE) +
  facet_grid(rows = vars(r_correlation),
             cols = vars(a_preferences), 
             labeller = label_both) +
  labs(x = "Mean(Rank)",
       y = "SD(Rank)",
       title = "Neighborhood Stability by Income",
       colour = "vision", linetype = "d_decay", shape = "d_decay") +
  theme_bw() +
  theme(legend.position = "bottom")

ggplot(stability_data) +
  aes(x = ave_rank_status,
      y = sd_rank_status,
      shape = as.factor(d_decay),
      linetype = as.factor(d_decay),
      colour = as.factor(vision)) +
  geom_smooth(se = FALSE) +
  facet_grid(rows = vars(r_correlation),
             cols = vars(a_preferences), 
             labeller = label_both) +
  labs(x = "Mean(Rank)",
       y = "SD(Rank)",
       title = "Neighborhood Stability by Status",
       colour = "vision", linetype = "d_decay", shape = "d_decay") +
  theme_bw() +
  theme(legend.position = "bottom")

ggplot(stability_data) +
  aes(x = ave_rank_rent,
      y = sd_rank_rent,
      shape = as.factor(d_decay),
      linetype = as.factor(d_decay),
      colour = as.factor(vision)) +
  geom_smooth(se = FALSE) +
  facet_grid(rows = vars(r_correlation),
             cols = vars(a_preferences), 
             labeller = label_both) +
  labs(x = "Mean(Rank)",
       y = "SD(Rank)",
       title = "Neighborhood Stability by Rent",
       colour = "vision", linetype = "d_decay", shape = "d_decay") +
  theme_bw() +
  theme(legend.position = "bottom")

ggplot(stability_data) +
  aes(x = ave_rank_quality,
      y = sd_rank_quality,
      shape = as.factor(d_decay),
      linetype = as.factor(d_decay),
      colour = as.factor(vision)) +
  geom_smooth(se = FALSE) +
  facet_grid(rows = vars(r_correlation),
             cols = vars(a_preferences), 
             labeller = label_both) +
  labs(x = "Mean(Rank)",
       y = "SD(Rank)",
       title = "Neighborhood Stability by Housing Quality",
       colour = "vision", linetype = "d_decay", shape = "d_decay") +
  theme_bw() +
  theme(legend.position = "bottom")
```
# Vacant Housing


```{r}
city_data %>%
  group_by(a_preferences, r_correlation, d_decay, vision) %>%
  summarise(vacancy_seg = mean(vacancy_seg)) %>%
  ggplot() +
  aes(x = vision,
      y = vacancy_seg,
      colour = as.factor(d_decay)) +
  geom_line() +
  facet_grid(rows = vars(r_correlation),
             cols = vars(a_preferences), 
             labeller = label_both) +
  labs(title = "Vacancy Segregation", y = "Theil Index H",
       colour = "d_decay") +
  theme_bw() +
  theme(legend.position = "bottom")

ggplot(nb_data) +
  aes(y = ave_vacant, 
      x = as.factor(vision),
      fill = as.factor(d_decay)) +
  geom_violin() +
  facet_grid(rows = vars(r_correlation),
             cols = vars(a_preferences), 
             labeller = label_both) +
  labs(y = "Average NB Vacancy Rate", fill = "d_decay", x = "vision") +
  theme_bw() +
  theme(legend.position = "bottom")
```


# Individual Level

## Rent

### Rent and Income

```{r}
ggplot(unit_data) +
  aes(y = rent, 
      x = as.factor(vision),
      fill = as.factor(d_decay)) +
  geom_violin() +
  facet_grid(rows = vars(r_correlation),
             cols = vars(a_preferences), 
             labeller = label_both) +
  labs(y = "Rent", fill = "d_decay", x = "Gini of the Income Distribution") +
  theme_bw() +
  theme(legend.position = "bottom")

ggplot(unit_data) +
  aes(y = log(rent_to_income), 
      x = as.factor(vision),
      fill = as.factor(d_decay)) +
  geom_violin() +
  geom_hline(yintercept = 1, 
             colour = "darkgrey") +
  facet_grid(rows = vars(r_correlation),
             cols = vars(a_preferences), 
             labeller = label_both) +
  labs(y = "Log(Rent / Income)", fill = "d_decay", x = "Gini of the Income Distribution") +
  theme_bw() +
  theme(legend.position = "bottom")

ggplot(unit_data) +
  aes(x = hh_income,
      y = rent,
      shape = as.factor(d_decay),
      linetype = as.factor(d_decay),
      colour = as.factor(vision)) +
  geom_smooth(se = FALSE) +
  facet_grid(rows = vars(r_correlation),
             cols = vars(a_preferences), 
             labeller = label_both) +
  labs(x = "Income",
       y = "Rent",
       colour = "vision", linetype = "d_decay", shape = "d_decay") +
  theme_bw() +
  theme(legend.position = "bottom")

ggplot(unit_data) +
  aes(x = hh_income,
      y = log(rent_to_income),
      shape = as.factor(d_decay),
      linetype = as.factor(d_decay),
      colour = as.factor(vision)) +
  geom_smooth(se = FALSE) +
  geom_hline(yintercept = 1, colour = "darkgrey") +
  facet_grid(rows = vars(r_correlation),
             cols = vars(a_preferences), 
             labeller = label_both) +
  labs(x = "Income",
       y = "Log(Rent / Income)",
       colour = "vision", linetype = "d_decay", shape = "d_decay") +
  theme_bw() +
  theme(legend.position = "bottom")

unit_data %>% 
  group_by(a_preferences, r_correlation, d_decay, vision) %>%
  summarise(prop = mean(rent > hh_income, na.rm = T)) %>%
  ggplot() +
  aes(x = vision, 
      y = prop,
      linetype = as.factor(d_decay)) +
  geom_point() +
  geom_line() +
  facet_grid(rows = vars(r_correlation),
             cols = vars(a_preferences), 
             labeller = label_both) +
  labs(x = "vision",
       y = "Proportion of rent-burdened Households",
       linetype = "d_decay", shape = "d_decay") +
  theme_bw() +
  theme(legend.position = "bottom")
```

## Housing Inequality

```{r}
ggplot(unit_data) +
  aes(y = housing_quality, 
      x = as.factor(vision),
      fill = as.factor(d_decay)) +
  geom_violin() +
  facet_grid(rows = vars(r_correlation),
             cols = vars(a_preferences), 
             labeller = label_both) +
  labs(y = "Housing Quality", fill = "d_decay", x = "Gini of the Income Distribution") +
  theme_bw() +
  theme(legend.position = "bottom")

ggplot(unit_data) +
  aes(x = hh_income,
      y = housing_quality,
      shape = as.factor(d_decay),
      linetype = as.factor(d_decay),
      colour = as.factor(vision)) +
  geom_smooth(se = FALSE) +
  facet_grid(rows = vars(r_correlation),
             cols = vars(a_preferences), 
             labeller = label_both) +
  labs(x = "Income",
       y = "Housing Quality",
       colour = "vision", linetype = "d_decay", shape = "d_decay") +
  theme_bw() +
  theme(legend.position = "bottom")

ggplot(unit_data) +
  aes(x = hh_status,
      y = housing_quality,
      shape = as.factor(d_decay),
      linetype = as.factor(d_decay),
      colour = as.factor(vision)) +
  geom_smooth(se = FALSE) +
  facet_grid(rows = vars(r_correlation),
             cols = vars(a_preferences), 
             labeller = label_both) +
  labs(x = "Status",
       y = "Housing Quality",
       colour = "vision", linetype = "d_decay", shape = "d_decay") +
  theme_bw() +
  theme(legend.position = "bottom")

reg_quality <- unit_data %>%
  nest_by(a_preferences, r_correlation, d_decay, as.factor(vision)) %>%
  mutate(trivariate = list(lm(data = data, housing_quality ~ hh_status + hh_income))) %>%
  reframe(tidy(trivariate)) %>%
  filter(r_correlation != 1 & term == "hh_status") %>%
  select(-statistic, -p.value) %>%
  rename(beta = estimate,
         se = std.error)

ggplot(reg_quality) +
  aes(x = as.factor(a_preferences),
      y = beta,
      ymin = beta - 3.1 * se,
      ymax = beta + 3.1 * se,
      colour = as.factor(vision)) +
  geom_errorbar(size = 2, width = 0, position = position_dodge(width = 0.5)) +
  geom_hline(yintercept = 0,
             colour = "darkgrey") +
  facet_grid(rows = vars(r_correlation),
             cols = vars(d_decay), 
             labeller = label_both) +
  labs(x = "a_preferences", y = "Coefficient of Status w. 99.9% CI", colour = "vision",
       title = "Linear Regression of Housing Quality on Status and Income") +
  theme_bw() +
  theme(legend.position = "bottom")
```


## Neighborhood Inequality

```{r}
ggplot(unit_data) +
  aes(y = ave_status, 
      x = as.factor(vision),
      fill = as.factor(d_decay)) +
  geom_violin() +
  facet_grid(rows = vars(r_correlation),
             cols = vars(a_preferences), 
             labeller = label_both) +
  labs(y = "Neighborhood Quality", fill = "d_decay", x = "Gini of the Income Distribution") +
  theme_bw() +
  theme(legend.position = "bottom")

ggplot(unit_data) +
  aes(x = hh_income,
      y = ave_status,
      shape = as.factor(d_decay),
      linetype = as.factor(d_decay),
      colour = as.factor(vision)) +
  geom_smooth(se = FALSE) +
  facet_grid(rows = vars(r_correlation),
             cols = vars(a_preferences), 
             labeller = label_both) +
  labs(x = "Income",
       y = "Neighborhood Quality",
       colour = "vision", linetype = "d_decay", shape = "d_decay") +
  theme_bw() +
  theme(legend.position = "bottom")

ggplot(unit_data) +
  aes(x = hh_status,
      y = ave_status,
      shape = as.factor(d_decay),
      linetype = as.factor(d_decay),
      colour = as.factor(vision)) +
  geom_smooth(se = FALSE) +
  facet_grid(rows = vars(r_correlation),
             cols = vars(a_preferences), 
             labeller = label_both) +
  labs(x = "Status",
       y = "Neighborhood Quality",
       colour = "vision", linetype = "d_decay", shape = "d_decay") +
  theme_bw() +
  theme(legend.position = "bottom")

reg_nbquality <- unit_data %>%
  nest_by(a_preferences, r_correlation, d_decay, as.factor(vision)) %>%
  mutate(trivariate = list(lm(data = data, ave_status ~ hh_status + hh_income))) %>%
  reframe(tidy(trivariate)) %>%
  filter(r_correlation != 1 & term == "hh_status") %>%
  select(-statistic, -p.value) %>%
  rename(beta = estimate,
         se = std.error)

ggplot(reg_nbquality) +
  aes(x = as.factor(a_preferences),
      y = beta,
      ymin = beta - 3.1 * se,
      ymax = beta + 3.1 * se,
      colour = as.factor(vision)) +
  geom_errorbar(size = 2, width = 0, position = position_dodge(width = 0.5)) +
  geom_hline(yintercept = 0,
             colour = "darkgrey") +
  facet_grid(rows = vars(r_correlation),
             cols = vars(d_decay), 
             labeller = label_both) +
  labs(x = "a_preferences", y = "Coefficient of Status w. 99.9% CI", colour = "vision",
       title = "Linear Regression of Neighborhood Quality on Status and Income") +
  theme_bw() +
  theme(legend.position = "bottom")
```

# Residential Mobility

```{r}
indiv_data <- unit_data %>%
  filter(!is.na(id)) %>%
  arrange(sample_id, r_correlation, a_preferences, d_decay, turnover, hh_id, t) %>%
  group_by(sample_id, r_correlation, a_preferences, d_decay, turnover, hh_id) %>%
  summarise(prob_move_unit = mean(obj_id != Lag(obj_id), na.rm = TRUE),
            prob_move_nb = mean(nb != Lag(nb), na.rm = TRUE),
            hh_income = mean(hh_income),
            hh_status = mean(hh_status))

ggplot(indiv_data) +
  aes(x = as.factor(vision),
      y = prob_move_unit,
      fill = as.factor(d_decay)) +
  geom_violin() +
  facet_grid(rows = vars(r_correlation),
             cols = vars(a_preferences), 
             labeller = label_both) +
  labs(x = "vision", y = "Probability of Moving", fill = "decay") +
  theme_bw() +
  theme(legend.position = "bottom")

ggplot(indiv_data) +
  aes(x = hh_income,
      y = prob_move_unit,
      linetype = as.factor(d_decay),
      colour = as.factor(vision)) +
  geom_smooth(se = FALSE) +
  facet_grid(rows = vars(r_correlation),
             cols = vars(a_preferences),
             labeller = label_both) +
  labs(x = "Income", y = "Probability of Moving", colour = "vision",
       shape = "d_decay", linetype = "d_decay") +
  theme_bw() +
  theme(legend.position = "bottom")

ggplot(indiv_data) +
  aes(x = hh_status,
      y = prob_move_unit,
      linetype = as.factor(d_decay),
      colour = as.factor(vision)) +
  geom_smooth(se = FALSE) +
  facet_grid(rows = vars(r_correlation),
             cols = vars(a_preferences),
             labeller = label_both) +
  labs(x = "Status", y = "Probability of Moving", colour = "vision",
       shape = "d_decay", linetype = "d_decay") +
  theme_bw() +
  theme(legend.position = "bottom")
```


# References

Fossett, Mark, and Warren Waren. 2005. “Overlooked Implications of Ethnic Preferences for Residential Segregation in Agent-Based Models.” Urban Studies 42(11):1893–1917. doi: 10.1080/00420980500280354.

Laurie, Alexander J., and Narendra K. Jaggi. 2003. “Role of ‘Vision’ in Neighbourhood Racial Segregation: A Variant of the Schelling Segregation Model.” Urban Studies 40(13):2687–2704. doi: 10.1080/0042098032000146849.

Lee, Barrett A., Sean F. Reardon, Glenn Firebaugh, Chad R. Farrell, Stephen A. Matthews, and David O’Sullivan. 2008. “Beyond the Census Tract: Patterns and Determinants of Racial Segregation at Multiple Geographic Scales.” American Sociological Review 73(5):766–91. doi:10.1177/000312240807300504.

Östh, John, William A. V. Clark, and Bo Malmberg. 2015. “Measuring the Scale of Segregation Using K-Nearest Neighbor Aggregates.” Geographical Analysis 47(1):34–49. doi:10.1111/gean.12053.

Reardon, Sean F., and Kendra Bischoff. 2011. “Income Inequality and Income Segregation.” American Journal of Sociology 116(4):1092–1153. doi: 10.1086/657114.

Reardon, Sean F., Stephen A. Matthews, David O’Sullivan, Barrett A. Lee, Glenn Firebaugh, Chad R. Farrell, and Kendra Bischoff. 2008. “The Geographic Scale of Metropolitan Racial Segregation.” Demography 45(3):489–514. doi:10.1353/dem.0.0019.
