---
title: "A socio-economic Model of Residential Segregation, Neighborhood Change and Housing Inequality"
subtitle: "Part IIIa: Analysis of the Main Experiment"
author: "Malte Grönemann"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, 
                      message = FALSE, 
                      warning = FALSE, 
                      error = FALSE,
                      out.width = "100%")

library(nanoparquet)
library(dplyr)
library(tidyr)
library(broom)
library(Hmisc)
library(DescTools)
library(ggplot2)
library(fixest)
library(modelsummary)

data_directory <- "./data/main/" # TODO change directory


## for some reason, loading the unit_data.parquet file created by the dataprep R script from the main experiment does not work.
## The file seems corrupted and every attempt at creating it new and saving it seems to fail as well.
## I therefore need to create the dataset again from the raw data for the analysis.
unit_data <- read_parquet(paste(data_directory,
                                "unit_raw.parquet", 
                                sep = "")) %>%
  filter(t >= 300)

parameters <- read_parquet(paste(data_directory, 
                                 "parameters.parquet", 
                                 sep = ""))

unit_data <- full_join(parameters, unit_data, 
                       by = "sample_id")

unit_data <- unit_data %>% 
  mutate(nb = paste(floor(x_coord / 5),
                    floor(y_coord / 5), 
                    sep = "_"),
         empty = is.na(hh_id),
         rent_to_income = rent / hh_income)

nb_data <- unit_data %>%
  group_by(sample_id, t, nb) %>%
  summarise(ave_inc = mean(hh_income, na.rm = TRUE),
            ave_status = mean(hh_status, na.rm = TRUE),
            ave_quality = mean(housing_quality, na.rm = TRUE),
            ave_rent = mean(rent, na.rm = TRUE),
            ave_vacant = mean(empty)) %>%
  mutate(rank_inc = length(unique(nb)) - rank(ave_inc) + 1, # inverting ranks to have the nb with highest ave as rank 1
         rank_status = length(unique(nb)) - rank(ave_status) + 1,
         rank_quality = length(unique(nb)) - rank(ave_quality) + 1,
         rank_rent = length(unique(nb)) - rank(ave_rent) + 1,
         rank_vacant = rank(ave_vacant)) # rank 1 is neighborhood with lowest vacancy rates

unit_data <- full_join(unit_data, nb_data, 
                       by = c("sample_id", "t", "nb"))


# loading uncorrupted files from storage
nb_data <- read_parquet(paste(data_directory, "nb_data.parquet", sep = ""))
city_data <- read_parquet(paste(data_directory, "city_data.parquet", sep = ""))
```

# Experiment Description

## Experiment Description

This main experiment simulates a 30x30 "city" (a toroidal grid) with 85% of all units being occupied. Therefore, there are 900 housing units and 765 households in each simulation. Income is distributed with a Beta(2, 5) distribution and 2 percent of all households are removed at each time step to model population dynamics and a corresponding number of newly created households is added to random empty housing units so that population density stays constant. Households calculate utility and landlords make their investment decisions based on the Moore neighborhood with a distance of one, so the 8 neighboring cells. 

The parameters to vary are the correlation between income and status, as well as the relative importance of housing quality versus neighbors in households' preferences. Both of these parameters vary from 0 to 1 in steps of 0.25, so have 5 levels. I also vary the decay parameter between 0.95, 0.9, 0.85 and 0.8. Each parameter combination is run 15 times, resulting in 5 x 5 x 4 x 15 = 1500 runs. Each run lasts 400 steps but only the last 100 steps are saved for analysis. The data are prepared and saved for analysis (see separate files) and reuse by other researchers. As the data record variables from every household and housing unit at every time step, I record 900 housing units x 1500 runs x 400 time steps = 540 million observations for housing units of which 135 million are analysed.

## Summary of Results


# City Level

## Segregation


```{r}
seg_long <- city_data %>%
  select(-ends_with("gini")) %>%
  pivot_longer(cols = ends_with("seg"),
               cols_vary = "slowest",
               names_to = "dimension",
               values_to = "segregation") %>%
  mutate(dimension = if_else(dimension == "inc_seg", "Income", dimension),
         dimension = if_else(dimension == "status_seg", "Status", dimension),
         dimension = if_else(dimension == "quality_seg", "Quality", dimension),
         dimension = if_else(dimension == "rent_seg", "Rent", dimension),
         dimension = if_else(dimension == "vacancy_seg", "Vacancy", dimension))

seg_long %>%
  filter(dimension != "Vacancy") %>%
  group_by(a_preferences, r_correlation, d_decay, dimension) %>%
  summarise(segregation = mean(segregation)) %>%
  ggplot() +
  aes(x = a_preferences,
      y = segregation,
      colour = dimension,
      group = dimension) +
  geom_point() +
  geom_line() +
  facet_grid(rows = vars(r_correlation),
             cols = vars(d_decay), 
             labeller = label_both) +
  labs(title = "Segregation", y = "H^R") +
  theme_bw() +
  theme(legend.position = "bottom")
```

In most circumstances, the model produces a considerable level of residential segregation by income, status as well as housing quality and rent. It is noteworthy that housing quality and rent are most unequally placed in space. Status is always least segregated and even not segregated (with some small index value due to random chance) when status is not correlated with income.

This is particularly surprising in the case where households only value neighbors’ status. Generally, with these preferences, segregation is lowest compared to the other preferences and lowest over all when income and status are uncorrelated. This can be interpreted as follows: when households want to live next to high status neighbors, they will move next to them if they can. Particularly rich households have more opportunities to move into desirable units as they can afford the desirable and therefore more expensive units. When rents in the neighborhood increase, landlords invest into housing quality, although it does not matter to the households in this case. But the high status households are not necessarily rich themselves, so they cannot afford staying in these desirable units and need to move away. This makes the neighborhood less attractive and the neighborhood they move to more attractive. The result is a “chase game” where the rich follow the popular and no stable segregated neighborhoods form. Results on neighborhood stability and residential mobility support this interpretation (see below). Only if status and income are correlated, high income and high status neighborhoods can form which prevent others from moving in via price and low income and low status households are forced to live next to each other. The “only status” column is very similar in its working to the model by Benard and Willer (2007) when prices are fully endogenous.

In contrast, when households only value housing quality and housing quality decays slowly, income and status segregation emerge purely as a consequence of the spatially interdependent decision rule of the landlords. And status segregation only emerges when income and status are correlated. Households sort into neighborhoods as a byproduct. Good quality housing units are the most desirable, therefore rich households will go for these and can price out other households. Because landlords react to changes in rent in the neighborhood, investing into housing quality is spatially interdependent. Therefore, distinct neighborhoods form where landlords invest into their housing quality, attracting rich households. On the other hand, neighborhoods losing affluent households will not invest and therefore housing quality declines. Poor households need to move to these units and neighborhoods as they cannot afford the better ones. Because housing quality has a lot of inertia, the neighborhoods with the high housing quality remain the neighborhoods with high housing quality and as a result the neighborhoods with high rents and high incomes. 

In cases where housing quality decays fast though, the NetLogo model (see robustness check E) suggests that segregation is low as even one period of disinvestment is enough for a unit to have lower utility than alternatives elsewhere and households move frequently, so that no stable neighborhoods form. Nevertheless, a preference for neighborhood status is not a necessary condition for social and economic segregation. Quite the opposite: segregation reaches its highest level in this condition. Because landlords invest based on rents in the local neighborhood, (not) investing into housing quality is contagious between landlords in a neighborhood. Landlords that do (not) invest into their housing stock increases (decreases) their rent, which motivates neighboring landlords to follow suit, further increasing (decreasing) rent in the neighborhood. There is again a self-reinforcing dynamic that stops when a marginal change in housing quality does not result in a meaningful change in rent anymore.

However, the assumption that landlords invest based on their neighborhood in the case where households do not value their neighborhood or do not value housing quality at all, is implausible.

The empirically most plausible case is when households value both housing quality and neighborhood status and income and status are correlated. In the fast decay condition, this is even the condition with the highest segregation indices as both mechanisms described above (spatial interdependence via (1) landlords investing in housing quality and (2) households seeking high status neighborhoods) reinforce and stabilize each other (this is even more apparent in robustness check E). The inferred dynamics fit well with the reviewed descriptions of neighborhood change: when a neighborhoods becomes more desirable as high status households move there or landlords invest in housing quality, rents rise and displace poor households. This frees space and additionally increases average status in the neighborhood as low income households tend to be low status. The new arrivals in the neighborhood then tend to be even richer and more desirable neighbors, and the cycle repeats until equilibrium when no new residents with sufficient income can be attracted anymore. Conversely, when a neighborhood looses residents with high incomes and/or status to other neighborhoods, rents in that neighborhood decrease. Landlords do not invest and housing quality decreases as well. The lower rents attract lower income households for whom it is still a good neighborhood though. Yet attractivity with lower status and no investments into housing quality may continue to fall. In equilibrium, a segregated state with a stable hierarchy of neighborhoods forms.

```{r}
modelsummary(list(
  feols(data = city_data,
              inc_seg ~  a_preferences + r_correlation + d_decay,
              cluster = "sample_id"),
  feols(data = city_data,
              inc_seg ~  a_preferences * r_correlation + d_decay,
              cluster = "sample_id"),
  feols(data = city_data,
              inc_seg ~  a_preferences * d_decay + r_correlation,
              cluster = "sample_id"),
  feols(data = city_data,
              inc_seg ~  a_preferences * r_correlation * d_decay,
              cluster = "sample_id")
),
  output = "html",
  stars = c('**' = .01, '***' = .001), 
  gof_map = c("nobs", "r.squared", "bic"))
```


Importantly, the two spatial interdependencies (households' preferences for neighbors and landlords investment decisions) reinforce and stabilize each other. If housing quality would decay fast, another neighborhood might provide better housing than ones current neighborhood after only a short time of disinvestment. And if households would only value neighborhood status, attractive neighborhoods with high-status but low-income residents can form, that attract more affluent households. These then displace the the very residents they moved to the neighborhood for. With only status as a preference and low correlation between income and status, a chase game can emerge where rich households follow high status households and segregation is low and neighborhoods are unstable. While stable segregation can occur with either exclusive preference (only status or only housing quality), it requires slowly decaying housing quality or a high correlation between income and desirability as a neighbor.



## Inequality

```{r}
ggplot(city_data) +
  aes(x = as.factor(a_preferences),
      y = rent_gini) +
  geom_violin() +
  facet_grid(rows = vars(r_correlation),
             cols = vars(d_decay), 
             labeller = label_both) +
  labs(title = "Rent Inequality",
       y = "Gini index", x = "a_preferences") +
  theme_bw() +
  theme(legend.position = "bottom")

ggplot(city_data) +
  aes(x = as.factor(a_preferences),
      y = quality_gini) +
  geom_violin() +
  facet_grid(rows = vars(r_correlation),
             cols = vars(d_decay), 
             labeller = label_both) +
  labs(title = "Inequality in Housing Quality",
       y = "Gini index", x = "a_preferences") +
  theme_bw() +
  theme(legend.position = "bottom")
```

# Neighborhood Level

## Income and Status

```{r}
ggplot(nb_data) +
  aes(y = ave_inc, 
      x = as.factor(d_decay)) +
  geom_violin() +
  facet_grid(rows = vars(r_correlation),
             cols = vars(a_preferences), 
             labeller = label_both) +
  labs(y = "Average NB Income", x = "d_decay") +
  theme_bw() +
  theme(legend.position = "bottom")

ggplot(nb_data) +
  aes(y = ave_status, 
      x = as.factor(a_preferences)) +
  geom_violin() +
  facet_grid(rows = vars(r_correlation),
             cols = vars(d_decay), 
             labeller = label_both) +
  labs(y = "Average NB Status", x = "a_preferences") +
  theme_bw() +
  theme(legend.position = "bottom")
```


## Rent and Housing Quality

```{r}
ggplot(nb_data) +
  aes(y = ave_rent, 
      x = as.factor(a_preferences)) +
  geom_violin() +
  facet_grid(rows = vars(r_correlation),
             cols = vars(d_decay), 
             labeller = label_both) +
  labs(y = "Average NB Rent", x = "a_preferences") +
  theme_bw() +
  theme(legend.position = "bottom")

ggplot(nb_data) +
  aes(y = ave_quality, 
      x = as.factor(a_preferences)) +
  geom_violin() +
  facet_grid(rows = vars(r_correlation),
             cols = vars(d_decay), 
             labeller = label_both) +
  labs(y = "Average NB Housing Quality", x = "a_preferences") +
  theme_bw() +
  theme(legend.position = "bottom")
```

## Neighborhood Stability

### Examples

```{r}
set.seed(53129)
nb_examples <- nb_data %>%
  filter(sample_id %in% sample(unique(nb_data$sample_id), 9))

ggplot(nb_examples) +
  aes(x = t, 
      y = ave_inc,
      colour = nb,
      group = nb) +
  geom_line() +
  facet_wrap(~sample_id) +
  labs(y = "Average NB Income") +
  theme(legend.position = "none")

ggplot(nb_examples) +
  aes(x = t, 
      y = ave_status,
      colour = nb,
      group = nb) +
  geom_line() +
  facet_wrap(~sample_id) +
  labs(y = "Average NB Status") +
  theme(legend.position = "none")

ggplot(nb_examples) +
  aes(x = t, 
      y = ave_rent,
      colour = nb,
      group = nb) +
  geom_line() +
  facet_wrap(~sample_id) +
  labs(y = "Average NB Rent") +
  theme(legend.position = "none")

ggplot(nb_examples) +
  aes(x = t, 
      y = ave_quality,
      colour = nb,
      group = nb) +
  geom_line() +
  facet_wrap(~sample_id) +
  labs(y = "Average NB Housing Quality") +
  theme(legend.position = "none")
```

### NB Averages over Time

```{r}
stability_data <- nb_data %>%
  group_by(sample_id, r_correlation, a_preferences, d_decay, nb) %>%
  summarise(ave_rank_inc = mean(rank_inc),
            sd_rank_inc = sd(rank_inc),
            ave_rank_status = mean(rank_status),
            sd_rank_status = sd(rank_status),
            ave_rank_quality = mean(rank_quality),
            sd_rank_quality = sd(rank_quality),
            ave_rank_rent = mean(rank_rent),
            sd_rank_rent = sd(rank_rent),
            ave_inc_t = mean(ave_inc, na.rm = TRUE),
            sd_inc_t = sd(ave_inc, na.rm = TRUE),
            ave_status_t = mean(ave_status, na.rm = TRUE),
            sd_status_t = sd(ave_status, na.rm = TRUE),
            ave_rent_t = mean(ave_rent, na.rm = TRUE),
            sd_rent_t = sd(ave_rent, na.rm = TRUE),
            ave_quality_t = mean(ave_quality, na.rm = TRUE),
            sd_quality_t = sd(ave_quality, na.rm = TRUE))


ggplot(stability_data) +
  aes(x = ave_inc_t,
      y = sd_inc_t,
      colour = as.factor(a_preferences)) +
  geom_smooth(se = FALSE) +
  facet_grid(rows = vars(r_correlation),
             cols = vars(d_decay), 
             labeller = label_both) +
  labs(x = "Mean",
       y = "SD",
       title = "Neighborhood Stability by Income",
       colour = "a_preferences") +
  theme_bw() +
  theme(legend.position = "bottom")

ggplot(stability_data) +
  aes(x = ave_status_t,
      y = sd_status_t,
      colour = as.factor(a_preferences)) +
  geom_smooth(se = FALSE) +
  facet_grid(rows = vars(r_correlation),
             cols = vars(d_decay), 
             labeller = label_both) +
  labs(x = "Mean",
       y = "SD",
       title = "Neighborhood Stability by Status",
       colour = "a_preferences") +
  theme_bw() +
  theme(legend.position = "bottom")

ggplot(stability_data) +
  aes(x = ave_rent_t,
      y = sd_rent_t,
      colour = as.factor(a_preferences)) +
  geom_smooth(se = FALSE) +
  facet_grid(rows = vars(r_correlation),
             cols = vars(d_decay), 
             labeller = label_both) +
  labs(x = "Mean",
       y = "SD",
       title = "Neighborhood Stability by Rent",
       colour = "a_preferences") +
  theme_bw() +
  theme(legend.position = "bottom")

ggplot(stability_data) +
  aes(x = ave_quality_t,
      y = sd_quality_t,
      colour = as.factor(a_preferences)) +
  geom_smooth(se = FALSE) +
  facet_grid(rows = vars(r_correlation),
             cols = vars(d_decay), 
             labeller = label_both) +
  labs(x = "Mean",
       y = "SD",
       title = "Neighborhood Stability by Housing Quality",
       colour = "a_preferences") +
  theme_bw() +
  theme(legend.position = "bottom")
```

### NB Rank over Time

```{r}
ggplot(stability_data) +
  aes(x = ave_rank_inc,
      y = sd_rank_inc,
      colour = as.factor(a_preferences)) +
  geom_smooth(se = FALSE) +
  facet_grid(rows = vars(r_correlation),
             cols = vars(d_decay), 
             labeller = label_both) +
  labs(x = "Mean(Rank)",
       y = "SD(Rank)",
       title = "Neighborhood Stability by Income Rank",
       colour = "a_preferences") +
  theme_bw() +
  theme(legend.position = "bottom")

ggplot(stability_data) +
  aes(x = ave_rank_status,
      y = sd_rank_status,
      colour = as.factor(a_preferences)) +
  geom_smooth(se = FALSE) +
  facet_grid(rows = vars(r_correlation),
             cols = vars(d_decay), 
             labeller = label_both) +
  labs(x = "Mean(Rank)",
       y = "SD(Rank)",
       title = "Neighborhood Stability by Status Rank",
       colour = "a_preferences") +
  theme_bw() +
  theme(legend.position = "bottom")

ggplot(stability_data) +
  aes(x = ave_rank_rent,
      y = sd_rank_rent,
      colour = as.factor(a_preferences)) +
  geom_smooth(se = FALSE) +
  facet_grid(rows = vars(r_correlation),
             cols = vars(d_decay), 
             labeller = label_both) +
  labs(x = "Mean(Rank)",
       y = "SD(Rank)",
       title = "Neighborhood Stability by Rent Rank",
       colour = "a_preferences") +
  theme_bw() +
  theme(legend.position = "bottom")

ggplot(stability_data) +
  aes(x = ave_rank_quality,
      y = sd_rank_quality,
      colour = as.factor(a_preferences)) +
  geom_smooth(se = FALSE) +
  facet_grid(rows = vars(r_correlation),
             cols = vars(d_decay), 
             labeller = label_both) +
  labs(x = "Mean(Rank)",
       y = "SD(Rank)",
       title = "Neighborhood Stability by Housing Quality Rank",
       colour = "a_preferences") +
  theme_bw() +
  theme(legend.position = "bottom")
```

# Vacant Housing


```{r}
city_data %>%
  group_by(a_preferences, r_correlation, d_decay) %>%
  summarise(vacancy_seg = mean(vacancy_seg)) %>%
  ggplot() +
  aes(x = a_preferences,
      y = vacancy_seg) +
  geom_line() +
  facet_grid(rows = vars(r_correlation),
             cols = vars(d_decay), 
             labeller = label_both) +
  labs(title = "Vacancy Segregation", y = "Theil Index H") +
  theme_bw() +
  theme(legend.position = "bottom")

ggplot(nb_data) +
  aes(y = ave_vacant, 
      x = as.factor(a_preferences)) +
  geom_violin() +
  facet_grid(rows = vars(r_correlation),
             cols = vars(d_decay), 
             labeller = label_both) +
  labs(y = "Average NB Vacancy Rate", x = "a_preferences") +
  theme_bw() +
  theme(legend.position = "bottom")
```

## Properties of Neighborhoods

```{r}
ggplot(nb_data) +
  aes(x = ave_inc,
      y = ave_vacant,
      colour = as.factor(a_preferences)) +
  geom_smooth(se = FALSE) +
  facet_grid(rows = vars(r_correlation),
             cols = vars(d_decay), 
             labeller = label_both) +
  labs(x = "Average NB Income", y = "Neighborhood Vacancy Rate", 
       colour = "a_preferences") +
  theme_bw() +
  theme(legend.position = "bottom")

ggplot(nb_data) +
  aes(x = ave_status,
      y = ave_vacant,
      colour = as.factor(a_preferences)) +
  geom_smooth(se = FALSE) +
  facet_grid(rows = vars(r_correlation),
             cols = vars(d_decay), 
             labeller = label_both) +
  labs(x = "Average NB Status", y = "Neighborhood Vacancy Rate", 
       colour = "a_preferences") +
  theme_bw() +
  theme(legend.position = "bottom")

ggplot(nb_data) +
  aes(x = ave_rent,
      y = ave_vacant,
      colour = as.factor(a_preferences)) +
  geom_smooth(se = FALSE) +
  facet_grid(rows = vars(r_correlation),
             cols = vars(d_decay), 
             labeller = label_both) +
  labs(x = "Average NB Rent", y = "Neighborhood Vacancy Rate", 
       colour = "a_preferences") +
  theme_bw() +
  theme(legend.position = "bottom")

ggplot(nb_data) +
  aes(x = ave_quality,
      y = ave_vacant,
      colour = as.factor(a_preferences)) +
  geom_smooth(se = FALSE) +
  facet_grid(rows = vars(r_correlation),
             cols = vars(d_decay), 
             labeller = label_both) +
  labs(x = "Ave. NB Housing Quality", y = "Neighborhood Vacancy Rate", 
       colour = "a_preferences") +
  theme_bw() +
  theme(legend.position = "bottom")
```

## Stability

```{r}
nb_data %>%
  group_by(sample_id, r_correlation, a_preferences, d_decay, nb) %>%
  summarise(ave_rank_vac = mean(rank_vacant),
            sd_rank_vac = sd(rank_vacant)) %>%
  ggplot() +
  aes(x = ave_rank_vac,
      y = sd_rank_vac,
      colour = as.factor(a_preferences)) +
  geom_smooth(se = FALSE) +
  facet_grid(rows = vars(r_correlation),
             cols = vars(d_decay), 
             labeller = label_both) +
  labs(x = "Average Neighborhood Rank by Vacancy Rate",
       y = "SD of Rank by Vacancy Rate", colour = "a_preferences") +
  theme_bw() +
  theme(legend.position = "bottom")
```

## Properties of empty Units

```{r}
unit_data %>%
  filter(empty) %>%
  ggplot() +
  aes(y = rent, 
      x = as.factor(a_preferences)) +
  geom_violin() +
  facet_grid(rows = vars(r_correlation),
             cols = vars(d_decay), 
             labeller = label_both) +
  labs(y = "Rent", x = "a_preferences") +
  theme_bw() +
  theme(legend.position = "bottom")

unit_data %>%
  filter(empty) %>%
  ggplot() +
  aes(y = housing_quality, 
      x = as.factor(a_preferences)) +
  geom_violin() +
  facet_grid(rows = vars(r_correlation),
             cols = vars(d_decay), 
             labeller = label_both) +
  labs(y = "Housing Quality", x = "a_preferences") +
  theme_bw() +
  theme(legend.position = "bottom")
```

# Residential Mobility

## Mobility between Units

```{r}
indiv_data <- unit_data %>%
  filter(!is.na(id)) %>%
  arrange(sample_id, r_correlation, a_preferences, d_decay, hh_id, t) %>%
  group_by(sample_id, r_correlation, a_preferences, d_decay, hh_id) %>%
  summarise(prob_move_unit = mean(obj_id != Lag(obj_id), na.rm = TRUE),
            prob_move_nb = mean(nb != Lag(nb), na.rm = TRUE),
            hh_income = mean(hh_income),
            hh_status = mean(hh_status))

ggplot(indiv_data) +
  aes(x = as.factor(a_preferences),
      y = prob_move_unit) +
  geom_violin() +
  facet_grid(rows = vars(r_correlation),
             cols = vars(d_decay), 
             labeller = label_both) +
  labs(x = "a_preferences", y = "Probability of Moving") +
  theme_bw() +
  theme(legend.position = "bottom")

ggplot(indiv_data) +
  aes(x = hh_income,
      y = prob_move_unit,
      colour = as.factor(a_preferences)) +
  geom_smooth(se = FALSE) +
  facet_grid(rows = vars(r_correlation),
             cols = vars(d_decay), 
             labeller = label_both) +
  labs(x = "Income", y = "Probability of Moving", colour = "a_preferences") +
  theme_bw() +
  theme(legend.position = "bottom")

ggplot(indiv_data) +
  aes(x = hh_status,
      y = prob_move_unit,
      colour = as.factor(a_preferences)) +
  geom_smooth(se = FALSE) +
  facet_grid(rows = vars(r_correlation),
             cols = vars(d_decay), 
             labeller = label_both) +
  labs(x = "Status", y = "Probability of Moving", colour = "a_preferences") +
  theme_bw() +
  theme(legend.position = "bottom")
```

## Mobility between Neighborhoods

```{r}
ggplot(indiv_data) +
  aes(x = as.factor(a_preferences),
      y = prob_move_nb) +
  geom_violin() +
  facet_grid(rows = vars(r_correlation),
             cols = vars(d_decay), 
             labeller = label_both) +
  labs(x = "a_preferences", y = "Probability of Moving NB") +
  theme_bw() +
  theme(legend.position = "bottom")

ggplot(indiv_data) +
  aes(x = hh_income,
      y = prob_move_nb,
      colour = as.factor(a_preferences)) +
  geom_smooth(se = FALSE) +
  facet_grid(rows = vars(r_correlation),
             cols = vars(d_decay), 
             labeller = label_both) +
  labs(x = "Income", y = "Probability of Moving NB", colour = "a_preferences") +
  theme_bw() +
  theme(legend.position = "bottom")

ggplot(indiv_data) +
  aes(x = hh_status,
      y = prob_move_nb,
      colour = as.factor(a_preferences)) +
  geom_smooth(se = FALSE) +
  facet_grid(rows = vars(r_correlation),
             cols = vars(d_decay), 
             labeller = label_both) +
  labs(x = "Status", y = "Probability of Moving NB", colour = "a_preferences") +
  theme_bw() +
  theme(legend.position = "bottom")
```

# Individual Level

## Rent

### Rent and Income

```{r}
ggplot(unit_data) +
  aes(y = rent, 
      x = as.factor(a_preferences)) +
  geom_violin() +
  facet_grid(rows = vars(r_correlation),
             cols = vars(d_decay), 
             labeller = label_both) +
  labs(y = "Rent", x = "a_preferences") +
  theme_bw() +
  theme(legend.position = "bottom")

ggplot(unit_data) +
  aes(y = log(rent_to_income), 
      x = as.factor(a_preferences)) +
  geom_violin() +
  geom_hline(yintercept = 1, 
             colour = "darkgrey") +
  facet_grid(rows = vars(r_correlation),
             cols = vars(d_decay), 
             labeller = label_both) +
  labs(y = "Log(Rent / Income)", x = "a_preferences") +
  theme_bw() +
  theme(legend.position = "bottom")

ggplot(unit_data) +
  aes(x = hh_income,
      y = rent,
      colour = as.factor(a_preferences)) +
  geom_smooth(se = FALSE) +
  facet_grid(rows = vars(r_correlation),
             cols = vars(d_decay), 
             labeller = label_both) +
  labs(x = "Income",
       y = "Rent",
       colour = "a_preferences") +
  theme_bw() +
  theme(legend.position = "bottom")

ggplot(unit_data) +
  aes(x = hh_income,
      y = log(rent_to_income),
      colour = as.factor(a_preferences)) +
  geom_smooth(se = FALSE) +
  geom_hline(yintercept = 1, colour = "darkgrey") +
  facet_grid(rows = vars(r_correlation),
             cols = vars(d_decay), 
             labeller = label_both) +
  labs(x = "Income",
       y = "Log(Rent / Income)",
       colour = "a_preferences") +
  theme_bw() +
  theme(legend.position = "bottom")
```

## Housing Inequality

```{r}
ggplot(unit_data) +
  aes(y = housing_quality, 
      x = as.factor(a_preferences)) +
  geom_violin() +
  facet_grid(rows = vars(r_correlation),
             cols = vars(d_decay), 
             labeller = label_both) +
  labs(y = "Housing Quality", x = "a_preferences") +
  theme_bw() +
  theme(legend.position = "bottom")

ggplot(unit_data) +
  aes(x = hh_income,
      y = housing_quality,
      colour = as.factor(a_preferences)) +
  geom_smooth(se = FALSE) +
  facet_grid(rows = vars(r_correlation),
             cols = vars(d_decay), 
             labeller = label_both) +
  labs(x = "Income",
       y = "Housing Quality",
       colour = "a_preferences") +
  theme_bw() +
  theme(legend.position = "bottom")

ggplot(unit_data) +
  aes(x = hh_status,
      y = housing_quality,
      colour = as.factor(a_preferences)) +
  geom_smooth(se = FALSE) +
  facet_grid(rows = vars(r_correlation),
             cols = vars(d_decay), 
             labeller = label_both) +
  labs(x = "Status",
       y = "Housing Quality",
       colour = "a_preferences") +
  theme_bw() +
  theme(legend.position = "bottom")

reg_quality <- unit_data %>%
  nest_by(a_preferences, r_correlation, d_decay) %>%
  mutate(trivariate = list(lm(data = data, housing_quality ~ hh_status + hh_income))) %>%
  reframe(tidy(trivariate)) %>%
  filter(r_correlation != 1 & term == "hh_status") %>%
  select(-statistic, -p.value) %>%
  rename(beta = estimate,
         se = std.error)

ggplot(reg_quality) +
  aes(x = as.factor(a_preferences),
      y = beta,
      ymin = beta - 3.1 * se,
      ymax = beta + 3.1 * se) +
  geom_errorbar(size = 2, width = 0, position = position_dodge(width = 0.5)) +
  geom_hline(yintercept = 0,
             colour = "darkgrey") +
  facet_grid(rows = vars(r_correlation),
             cols = vars(d_decay), 
             labeller = label_both) +
  labs(x = "a_preferences", y = "Coefficient of Status w. 99.9% CI",
       title = "Linear Regression of Housing Quality on Status and Income") +
  theme_bw() +
  theme(legend.position = "bottom")
```

## Neighborhood Inequality

```{r}
ggplot(unit_data) +
  aes(y = ave_status, 
      x = as.factor(a_preferences)) +
  geom_violin() +
  facet_grid(rows = vars(r_correlation),
             cols = vars(d_decay), 
             labeller = label_both) +
  labs(y = "Neighborhood Quality", x = "a_preferences") +
  theme_bw() +
  theme(legend.position = "bottom")

ggplot(unit_data) +
  aes(x = hh_income,
      y = ave_status,
      colour = as.factor(a_preferences)) +
  geom_smooth(se = FALSE) +
  facet_grid(rows = vars(r_correlation),
             cols = vars(d_decay), 
             labeller = label_both) +
  labs(x = "Income",
       y = "Neighborhood Quality",
       colour = "a_preferences") +
  theme_bw() +
  theme(legend.position = "bottom")

ggplot(unit_data) +
  aes(x = hh_status,
      y = ave_status,
      colour = as.factor(a_preferences)) +
  geom_smooth(se = FALSE) +
  facet_grid(rows = vars(r_correlation),
             cols = vars(d_decay), 
             labeller = label_both) +
  labs(x = "Status",
       y = "Neighborhood Quality",
       colour = "a_preferences") +
  theme_bw() +
  theme(legend.position = "bottom")

reg_nbquality <- unit_data %>%
  nest_by(a_preferences, r_correlation, d_decay) %>%
  mutate(trivariate = list(lm(data = data, ave_status ~ hh_status + hh_income))) %>%
  reframe(tidy(trivariate)) %>%
  filter(r_correlation != 1 & term == "hh_status") %>%
  select(-statistic, -p.value) %>%
  rename(beta = estimate,
         se = std.error)

ggplot(reg_nbquality) +
  aes(x = as.factor(a_preferences),
      y = beta,
      ymin = beta - 3.1 * se,
      ymax = beta + 3.1 * se) +
  geom_errorbar(size = 2, width = 0, position = position_dodge(width = 0.5)) +
  geom_hline(yintercept = 0,
             colour = "darkgrey") +
  facet_grid(rows = vars(r_correlation),
             cols = vars(d_decay), 
             labeller = label_both) +
  labs(x = "a_preferences", y = "Coefficient of Status w. 99.9% CI",
       title = "Linear Regression of Neighborhood Quality on Status and Income") +
  theme_bw() +
  theme(legend.position = "bottom")
```

## Residential Mobility

```{r}
indiv_data <- unit_data %>%
  filter(!is.na(id)) %>%
  arrange(sample_id, r_correlation, a_preferences, d_decay, hh_id, t) %>%
  group_by(sample_id, r_correlation, a_preferences, d_decay, hh_id) %>%
  summarise(prob_move_unit = mean(obj_id != Lag(obj_id), na.rm = TRUE),
            prob_move_nb = mean(nb != Lag(nb), na.rm = TRUE),
            hh_income = mean(hh_income),
            hh_status = mean(hh_status))

ggplot(indiv_data) +
  aes(y = prob_move_unit,
      x = as.factor(a_preferences)) +
  geom_violin() +
  facet_grid(rows = vars(r_correlation),
             cols = vars(d_decay), 
             labeller = label_both) +
  labs(y = "Probability of Moving", x = "a_preferences") +
  theme_bw() +
  theme(legend.position = "bottom")

ggplot(indiv_data) +
  aes(x = hh_income,
      y = prob_move_unit,
      colour = as.factor(a_preferences)) +
  geom_smooth(se = FALSE) +
  facet_grid(rows = vars(r_correlation),
             cols = vars(d_decay), 
             labeller = label_both) +
  labs(x = "Income", y = "Probability of Moving", colour = "a_preferences") +
  theme_bw() +
  theme(legend.position = "bottom")

ggplot(indiv_data) +
  aes(x = hh_status,
      y = prob_move_unit,
      colour = as.factor(a_preferences)) +
  geom_smooth(se = FALSE) +
  facet_grid(rows = vars(r_correlation),
             cols = vars(d_decay), 
             labeller = label_both) +
  labs(x = "Status", y = "Probability of Moving", colour = "a_preferences") +
  theme_bw() +
  theme(legend.position = "bottom")
```
